---
title:  "Computational Practical 1"
author: "[Matthew Aldridge](mailto:m.aldridge@leeds.ac.uk)"
date:   "MATH2750 Introduction to Markov Processes (2019--20)"
output: html_document  
---
<!-- or "output: pdf_document" to make a LaTeXed PDF -->



*This computer practical is an oppotunity to learn more about Markov chains using R. This practical not an assessed part of the course that counts towards your mark, but Computer Practical 2 will be assessed, so this will be useful practice. The material in this practical* is *examinable.*

*You should work though this problem sheet in the practical session for this course, between Monday 17 and Thursday 20 February. Check your timetable (look for week 17) to find the time and room for your session. (The webpage <https://leeds.service-now.com/it?id=clusters> will give you directions.) A member of staff will be present to help you with any difficulties. There may not be enough time to do every question -- if you manage to complete Questions 0, 1 and 2, you're making good progress; if you complete all of Questions 0 to 4, then you are doing very well. You can, of course, continue working on the sheet in your own time after the practical session.*

*This question sheet is available in two formats:*

* *[`C1.html`](https://mpaldridge.github.io/math2750/C1.html) is an easy-to-read HTML file of the questions and comments. You should open this in a web browser.*
* *[`C1.Rmd`](https://mpaldridge.github.io/math2750/C1.Rmd) is a plain text file. I recommend downloading this file and opening it in RStudio. This will make it easy to run the R code included here, by clicking the green "play" button by each chunk of R code. (This file is written in a language called "[R Markdown](https://www.stat.cmu.edu/~cshalizi/rmarkdown/)", which you could choose to use for writing your report.)*

*This question sheet assumes you can open RStudio (or operate R in whatever way you prefer), and that you can use basic commands such as `x <- c(1, 2, 3)` and `mean(x)`. The University of Leeds* [Introduction to R](http://www1.maths.leeds.ac.uk/statistics/R/Rintro.pdf) *covers more basic topics.*



***



**Question 0:** *Answer as many of the questions on this problem sheet as you can. Write up a short report on your findings, with answers to the questions you have studied. Make sure to include relevant graphs produced and the R code you used.*


If you would like some feedback on your report, you can email it to <m.aldridge@leeds.ac.uk> or hand a paper copy to me in lectures. I will provide feedback on any reports received by Thursday 27 February. I am happy for your report to be submitted in any format, including: HTML or PDF produced by [R Markdown](https://www.stat.cmu.edu/~cshalizi/rmarkdown/), PDF produced by LaTeX, Microsoft Word document, or handwritten report with R printouts for plots.



**Question 1.** *We will be studying the simple random walk starting from $X_0 = 0$ with up probability $p = 0.6$ and down probability $q = 1 - p = 0.4$. Produce a sample of the first $N = 10$ steps of this simple random walk.*


We want to start by building the increments process $(Z_n)$, where each $Z_n$ is $+1$ with probability $p = 0.6$ and $-1$ with probability $q = 0.4$. We will use the `sample` function to do this. You can find out how to use `sample` by reading the help file, which you can open by entering `?sample` in the console. What arguments do you need to give to `sample` to get $10$ values of $\pm 1$ with the correct probabilities? What, for example, do we get from the following?

<!-- Delete `eval = FALSE` if you want to run the R code in a R Markdown document. -->
```{r, eval = FALSE}
Z <- sample(c(1, 7, 4), 20, replace = TRUE, prob = c(0.5, 0.3, 0.2))
Z
```

What is the purpose of the argument `replace = TRUE`? Can you adapt the above code to get the desired increments process?

Once we have the increments process $(Z_n)$, we need to transform it into a random walk $(X_n)$. A useful function here is the *cumulative sum* function `cumsum`. The cumulative sum of a vector $(y_1, y_2, \dots, y_n)$ is the vector of partial sums
$$ (y_1, y_1 + y_2, y_1 + y_2 + y_3, \dots, y_1 + y_2 + \cdots + y_n) . $$
How can we use this function to form a random walk from the increments?

All together, your code to produce the random walk should look something like this:

```{r, eval = FALSE}
p <- 0.6
q <- 1 - p
N <- 10
Z <- # Add your code for increments
Z    # Should produce a string of 10 plus-or-minus 1s
X <- # Turn increments into a random walk
X    # Should produce something that looks like a random walk
```



**Question 2.** *Plot a sample of the random walk for $N = 10$ steps and for $N = 10000$ steps. Try to make your plots look smart, for example by giving them titles and labelling the axes. Briefly comment on the differences between the plots.*


The standard command to plot a vector $\mathbf{X}$ of length $N$ against $1, 2, \dots, N$ is

```{r, eval = FALSE} 
plot(1:N, X)  # plot(X) is a shortcut for this
```

Here `1:N` means the vector $(1, 2, \dots, N)$.

You will probably want to include the starting point $X_0 = 0$ on your plot too. How can you do this?

The plot function can take extra optional arguments. You can find out about these by typing `?plot`, which brings up the help for the function `plot`. For example, what does the following do?

```{r, eval = FALSE} 
plot(1:N, X,
     type = "b",
     col  = "red",
     ylim = c(-15, 15),
     xlab = "Test",
     main = "Hello!")
```

Use `?plot` (or Google, or ask a friend) to find out the possible arguments to `type = `, and make sure to pick the most appropriate one. You may find that different arguments are appropriate for the $N = 10$ and $N = 10000$ cases.

Remember to comment on the differences between the plots.



**Question 3.** *Make a function in R that produces a sample of a simple random walk for given $p$ of length $N$. Test your function to check it works. What are the advantages of making a a function like this?*


You'll want a syntax something like this:

```{r, eval = FALSE}
RandomWalk <- function(N, p) {
  Z <- # Your code for N increments that are +1 with probability p
       # and -1 otherwise
  cumsum(Z)
}
```

If you do this correctly, then, for example,

```{r, eval = FALSE}
X <- RandomWalk(20, 0.3)
X
```

should produce a simple random walk of length $20$ with $p = 0.3$. It might be helpful to check by plotting a random walk of your choice.



**Question 4.** *Estimate the expected value of of the simple random walk at $N = 100$ steps by simulating many random walks and taking an average. How does this compare with the true answer?*


Given a vector `x`, we know that, for example, `x[10]` gives the 10th element of the vector. So, using your `RandomWalk` function, `RandomWalk(N, p)[N]` should give the final value of an random walk of length $N$.
  
We can use the `replicate` function (read `?replicate` to learn more) to get a large number of samples of the random walk.

```{r, eval = FALSE}
N <- 100
trials <- 10000  # Or some other appropriate large number
samples <- replicate(trials, RandomWalk(N, p)[N])
# Now add code to output the average of samples
```

What does theory we have learned say the answer should be?



**Question 5.** (Optional) *Investigate estimating other properties of the simple random walk through simulation.*


